---
title: "Gravitational Wave Analysis in Python"
subtitle: "CSE 515: Software Design For Data Science"
author: "Hannah Luebbering"
output:
  html_document: 
    css: assets/main.css
    toc: yes
    toc_float: yes
    self_contained: yes
bibliography: assets/references.bib
nocite: '@*'
# knit: pagedown::chrome_print
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = ".") })
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(kableExtra)
library(dplyr)
library(knitr)
library(utils)

```


<script src="assets/pymin.js"></script>
<script>hljs.initHighlightingOnLoad();</script>


## Introduction


> **GravitationalWaves:** a Python package designed to simulate, detect, and analyze continuous gravitational wave-forms. The package also takes in observed data for comparison in detection and analysis.


This report details through the software design decisions, usage guidelines, and extensibility of the GravitationalWaves package system, with respect to existing software libraries. The software libraries used for the basis of this project include `legwork` and `riroriro`. Legwork is a python package for computing evolution and detection rates of gravitational-wave sources discovered by space-based detectors such as LISA. Riroriro is a python package for simulating and evaluating gravitational waves.




### Background


Gravitational waves, first theorized by Einstein, arise during events such as the collision of supermassive binary black hole mergers. A binary black hole is a system consisting of two black holes in close orbit around each other. These collisions are so powerful that they create distortions in space time, known as gravitational waves.


**Definition.** Gravitational waves are invisible distortions in space time, caused by massive events such as collisions between two black holes or neutron stars.


The Laser Interferometer Space Antenna is a proposed space probe to detect and accurately measure gravitational waves—tiny ripples in the fabric of spacetime—from astronomical sources. The space-based gravitational wave detector LISA presents a new view of gravitational waves by focusing on lower frequencies $\left( 10^{−5} \lt f / \mathrm{Hz} \lt 10^{-1} \right)$ than ground-based detectors. This enables the study of many new source classes including mergers of supermassive black holes, extreme mass ratio inspirals, and cosmological GW backgrounds. 


The importance of studying gravitational wave-forms stems from the idea of detecting and using gravity to estimate other dynamical astrophysical phenomena, giving enormous potential for discovering parts of the universe that are invisible to the eye, such as black holes and other unknowns. 


\ 



### The Package


`GravitationalWaves` is a Python package for simulating the gravitational wave-forms of binary mergers of black holes and neutron stars, computing several properties of these mergers and wave-forms, and evaluating their detectability. In addition, the package also takes in observed data from gravitational wave detectors, such as the Laser Interferometer Space Antenna (LISA), to compare data and predict detection rates for stellar-origin binaries. Specifically, we implement the expressions for the evolution of binary orbits due to the emission of gravitational waves, equations for the strain amplitudes and signal-to-noise ratios of binaries, and approximations for the LISA sensitivity curves.



Similar to the existing software libraries of `legwork` and `riroriro`, the GravitationalWaves package represents an implementation of gravitational wave tools, focusing on a mathematical approach for analyzing different sources. Just like legwork, this package provides mechanisms to calculate orbits strain amplitudes, compute sensitivity curves, and calculate signal to noise ratios of a collection of gravitational wave sources. Furthermore, the package provides tools and plotting functions for visualizing such results. Just like riroriro, this package provides gravitational waveform simulation that's based on a computational implementation of an earlier theoretical waveform model that uses post-Newtonian expansions and simplified Einstein field equations.



#### Package Data



GravitationalWaves uses simulated data for gravitational waves and observations from detectors such as the Laser Interferometer Space Antenna (LISA).

**Simulated Data:** We obtain the simulated data by implementing the package functions for constructing simulations of gravitational waveform signals from a binary merger of two black holes or two neutron stars and outputting the data of such signals in terms of frequency and strain amplitude. 

```python
i_amp = ins.inspiral_strain_amplitude(i_Aorth, i_Adiag)
i_freq = mat.frequency_SI_units(i_m_omega, M)
```


```{r, fig.cap="Simulated Data Example"}
simulated.df <- data.frame("Amp" = c("7.9097e-22", "7.9105e-22", "7.9112e-22", "7.9121e-22", "7.9128e-22", "7.9136e-22", "7.9144e-22"), "Freq" = c(10.0066, 10.0081, 10.0096, 10.0111, 10.0126, 10.0141, 10.0156))

kable(head(simulated.df, 4), col.names = c("Strain Amplitude", "Frequency"), escape = FALSE) %>%
  kable_styling(bootstrap_options = c("hover", "striped"), full_width = F, html_font = "Roboto", font_size = 11) %>%
  row_spec(0:4, hline_after = F, extra_css = "border-bottom: 1px solid #FEFEFE;") %>%
  column_spec(1, color = "#111111", extra_css = "font-weight: 400;") %>%
  column_spec(1:2, color = "#555", extra_css = "font-weight: 300; text-align: right;") %>%
  row_spec(0, color = "#111111", extra_css = "text-transform: uppercase; letter-spacing: 1.25px; font-weight: 400; font-size: 10px")
```


\ 


**Observed Data:** The observed data contains LISA-validated binaries obtained from the Kupfer et al. 2018 study. These observations are preloaded to the package and used to help generate the Source class based on these provided binaries.



```{r}
table1 <- read.csv("assets/data/table1.csv")
table2 <- read.csv("assets/data/table2.csv")
colnames(table1) <- c("Source", "$I_\\mathrm{Gal}$", "$b_\\mathrm{Gal}$", "Orbital Period", "$m_1$", "$m_2$", "t", "Refs", "Type")
colnames(table2) <- c("Source", "f (mHz)", "$\\beta$ (mas)", "$\\sigma_\\beta$ (mas)", "d (pc)", "$\\sigma_d$ (pc)", "$\\mathcal{A}$", "SNR", "type")
```


```{r}
tab1 <- kable(table1, escape = FALSE) %>%
  remove_column(columns = 9) %>%
  kableExtra::group_rows(start_row = 1, end_row = 11, group_label = "AM CVn type") %>%
  kableExtra::group_rows(start_row = 12, end_row = 15, group_label = "Detached white dwarfs") %>%
  kableExtra::group_rows(start_row = 16, end_row = 16, group_label = "Hot subdwarf binaries") %>%
  kable_styling(bootstrap_options = c("hover", "striped"), full_width = F, html_font = "Roboto", font_size = 10.5) %>%
  row_spec(0:18, hline_after = F, extra_css = "border-bottom: 1px solid #FEFEFE;") %>%
  column_spec(1, color = "#111111", extra_css = "font-weight: 400;") %>%
  column_spec(2:8, color = "#555", extra_css = "font-weight: 300; font-family: Roboto Condensed !important;") %>%
  row_spec(0, color = "#111111", extra_css = "text-transform: uppercase; letter-spacing: 1.25px; font-weight: 400; font-size: 10px; font-family: Roboto;")
  
```



```{r}
kable(table2, escape = FALSE) %>% 
  remove_column(columns = 9) %>%
  kableExtra::group_rows(start_row = 1, end_row = 11, group_label = "AM CVn type") %>%
  kableExtra::group_rows(start_row = 12, end_row = 15, group_label = "Detached white dwarfs") %>%
  kableExtra::group_rows(start_row = 16, end_row = 16, group_label = "Hot subdwarf binaries") %>%
  kable_styling(bootstrap_options = c("hover", "striped", "condensed"), full_width = F, html_font = "Roboto", font_size = 10.5) %>%
  row_spec(0:18, hline_after = F, extra_css = "border-bottom: 1px solid #FEFEFE;") %>%
  column_spec(1, color = "#111111", extra_css = "font-weight: 400;") %>%
  column_spec(2:8, color = "#555", extra_css = "font-weight: 300; font-family: Roboto Condensed !important;") %>%
  row_spec(0, color = "#111111", extra_css = "text-transform: uppercase; letter-spacing: 1.25px; font-weight: 400; font-size: 10px; font-family: Roboto;")
```


\ 



**Limitations:** Both simulated and observed calculations apply a low-order post-Newtonian description of gravitational wave emission, meaning that the data do not account for higher-order effects.


\ 




#### Software Requirements

The following table presents the fundamental package dependencies:
 


```{r}
dependencies <- data.frame("Package" = c("python", "pip", "numba", "numpy", "astropy", "scipy", "matplotlib", "seaborn", "schwimm", "legwork"), "Version" = c("$\\geq$ 3.7.0", "$\\geq$ 21.0.0", "$\\geq$ 0.5.0", "$\\geq$ 1.16.0", "$\\geq$ 4.0.0", "$\\geq$ 1.5.0", "$\\geq$ 3.3.2", "$\\geq$ 0.11.1", "$\\geq$ 0.3.2", "$\\geq$ 0.2.4"), 
                           "Type" = c("Programming language", "Python package", "numba", "numpy", "astropy", "scipy", "matplotlib", "seaborn", "schwimm", "legwork"),
                           "Description" = c("High-level, general-purpose programming language", "Recommended package installer for Python", "NumPy-aware optimizing compiler using LLVM", "Fundamental package for array computing", "Astronomy and astrophysics core library", "Software for mathematics, science, and engineering", "Library for creating various visualizations in Python", "Library for making statistical graphics in Python", "Common interface for parallel processing pools", "Package for analyzing gravitational waves"))

kable(dependencies, col.names = c("Package", "Version", "Type", "Description"), escape = FALSE) %>%
  kable_styling(bootstrap_options = c("hover", "striped", "condensed"), full_width = T, html_font = "Roboto", font_size = 11) %>%
  row_spec(0:10, hline_after = F, extra_css = "border-bottom: 1px solid #FEFEFE;") %>%
  column_spec(1:3, color = "#555", extra_css = "font-weight: 300; text-align: left;") %>%
  row_spec(0, color = "#111111", extra_css = "text-transform: uppercase; letter-spacing: 1.25px; font-weight: 400; font-size: 10px; text-align: left;") %>%
    column_spec(1, color = "#111111", extra_css = "font-weight: 400; text-align: right;") %>%
    column_spec(2, color = "#111111", extra_css = "font-size: 11px;") %>%
    column_spec(4, color = "#111111", extra_css = "font-size: 10px; text-align: justify; font-weight: 100 !important;", width = "8cm")
```


\ 


#### Project Structure




The structure of the GravitationalWaves github repository is as follows:



<div class = "roundedlist2">
<li class="folder"> 
<a>GravitationalWaves </a>
<ul class="directory-list">
  <li class="folder">
    <a>docs </a> <span class= "infolder">(folder contains functional and design specification)</span>
    <ul>
      <li>presentation.pdf</li>
    </ul>
  </li>
  <li class="folder">
    <a>examples </a>
    <ul>
      <li>01_InstantiateSourceClass.ipynb</li>
      <li>02_CalculateSNR.ipynb</li>
      <li>03_PlotSourceDistribution.ipynb</li>
    </ul>
  </li>
  <li class="folder">
    <a>GravitationalWaves </a> <span class= "infolder">(python package folder)</span>
    <ul>
      <li class="folder">
        <a>tests </a>
        <ul>
          <li>test_psd.py</li>
          <li>test_source.py</li>
          <li>test_strain.py</li>
          <li>test_utils.py</li>
          <li>test_visualization.py</li>
          <li>test_wavesim.py</li>
          <li>__init__.py</li>
        </ul>
      </li>
      <li>psd.py</li>
      <li>source.py</li>
      <li>strain.py</li>
      <li>utils.py</li>
      <li>visualization.py</li>
      <li>wavesim.py</li>
      <li>__init__.py</li>
    </ul>
  </li>
  <li>LICENSE <span class= "infolder">(open source MIT License)</span> </li>
  <li>README.md <span class= "infolder">(overview of the project)</span> </li>
  <li>setup.py <span class= "infolder">(file initializes project after it has been cloned)</span> </li>
</ul>
</li>

</div>







From the above, we can see the top-level package called `GravitationalWaves`, with six modules: `strain.py`, `psd.py`, `utils.py`, `visualization.py`, `source.py`, `wavesim.py`. There's also the subpackage called `tests` located in the top-level package. There's also an examples and docs directory that contains demos and documentation of the package, though, they're not packages since both don't include an `__init__.py`.






--------------------------------------------



## Software Components 


The software components for the `GravitationalWaves` package include the package's modules, classes, and functions. A Python package is any directory with an `__init__.py` file, and this file gathers all package-wide definitions to import every module in the package. Python modules are an essential abstraction layer, which allow separating code into parts holding related data and functionality. For example, a layer of a project might handle interfacing with user actions, while another handles data manipulation. To separate these two layers, we'd regroup all user interfacing functionality into one file and all data manipulation functionality into another file.




### Component Specification



For the following component specification, we define each components' metadata, interface, and implementation. The metadata consists of a component's name and description. The interface specifies the list of inputs/outputs and their properties. And the implementation specifies how to execute the component instance, pass inputs to the component code, and get the component's outputs. After the component specification, I introduce the component code, which implements the logic needed to perform a step in the package workflow.


\ 



#### Module 1. source.py

<div class = "methods1">

*Name:* Source (class)

*Description:* The source module provides a simple interface to the functions in the remaining modules. This module contains the Source class for analyzing a generic set of gravitational wave sources, stationary/evolving and circular/eccentric.

*Annotations:*

    """
    Class for generic gravitational wave (GW) sources.
    """
            
*Inputs:*

- {**name:** `m_1`, **description:** `Primary mass`, **type:** `Int`, **optional:** `False`}
- {**name:** `m_2`, **description:** `Secondary mass`, **type:** `Int`, **optional:** `False`}
- {**name:** `ecc`, **description:** `Initial eccentricity`, **type:** `Int`, **optional:** `False`}
- {**name:** `dist`, **description:** `Luminosity distance to source`, **type:** `Int`, **optional:** `False`}
- {**name:** `n_proc`, **description:** `Number of processors to split`, **type:** `Int`, **default:** `1`, **optional:** `True`}
- {**name:** `f_orb`, **description:** `Orbital frequency`, **type:** `Int`, **optional:** `False`}
- {**name:** `a`, **description:** `Semi-major axis`, **type:** `Int`, **default:** `None`, **optional:** `True`}
- {**name:** `position`, **description:** `Sky position of source`, **type:** `Int`, **default:** `None`, **optional:** `True`}
- {**name:** `weights`, **description:** `Statistic weights for each sample`, **type:** `Int`, **default:** `None`, **optional:** `True`}
- {**name:** `gw_lum_tol`, **description:** `Allowed error on GW luminosity`, **type:** `Int`, **default:** `0.05`, **optional:** `True`}
- {**name:** `stat_tol_e`, **description:** `Frequency change above which a binary is stationary`, **type:** `Int`, **default:** `1e-2`, **optional:** `True`}
- {**name:** `interpolate_g`, **description:** `Whether to interpolate g(n,e)`, **type:** `Int`, **default:** `True`, **optional:** `True`}
- {**name:** `interpolate_sc`, **description:** `Whether to interpolate LISA sensitivity curve`, **type:** `Int`, **default:** `True`, **optional:** `True`}

*Outputs:*

- {**name:** `snr`, **type:** `Float`, **description:** `Signal-to-noise ratio`}


*Implementation:*

```python
import GravitationalWaves.source 
import astropy.units as u

source1 = source.Source(m_1 = 11 * u.Msun, m_2 = 11 * u.Msun, ecc = 0.3, dist = 9 * u.kpc, f_orb = 1e-4 * u.Hz, interpolate_g = False)
```

</div>




----------------------------------------



#### Module 2. strain.py

<div class = "methods1">

*Name:* strain

*Description:* A program that uses functions from `legwork` to calculate signal-to-noise ratio of sources based on four different cases. For the different cases, the module contains four functions that cover whether a source is circular-stationary, circular-evolving, eccentric-stationary, or eccentric-evolving.

*Annotations:*

    """
    Computes strain amplitude at `n` harmonics
    """
  
*Inputs:*

- {**name:** `m_c`, **description:** `Chirp mass of each binary`, **type:** `Float`, **optional:** `False`}
- {**name:** `f_orb`, **description:** `Orbital frequency of each binary`, **type:** `Float`, **optional:** `False`}
- {**name:** `ecc`, **description:** `Eccentricity of each binary`, **type:** `Float`, **optional:** `False`}
- {**name:** `n`, **description:** `Harmonic(s) to calculate the strain`, **type:** `Int`, **optional:** `False`}
- {**name:** `dist`, **description:** `Distance to each binary`, **type:** `Float`, **optional:** `False`}
- {**name:** `position`, **description:** `Sky position of source`, **default:** `None`, **type:** `Array`, **optional:** `True`}
- {**name:** `interpolated_g`, **description:** `G(n,e) from Peters (1964)`, **default:** `None`, **type:** Function, **optional:** `True`}
         
*Outputs:*

- {**name:** `h_0`, **type:** `Float`, **description:** `Strain amplitude`}

*Implementation:*


```python
from GravitationalWaves import strain, source
import numpy as np
import astropy.units as u

# create a random collection of sources
n_values = 1500
mass1 = np.random.uniform(0, 10, n_values) * u.Msun
mass2 = np.random.uniform(0, 10, n_values) * u.Msun
lumDist = np.random.normal(8, 1.5, n_values) * u.kpc
freqOrb = 10**(-5 * np.random.power(3, n_values)) * u.Hz
ecc = 1 - np.random.power(3, n_values)
sources = source.Source(m_1=m_1, m_2=m_2, ecc=ecc, dist=dist, f_orb=f_orb, gw_lum_tol=0.05, stat_tol=1e-2,
                        interpolate_g=True, interpolate_sc=True})
                        
h_c_n4 = sources.get_h_c_n(harmonics=[1, 2, 3, 4])
```

</div>



----------------------------------------




#### Module 3. psd.py


<div class = "methods1">

*Name:* psd

*Description:* The psd module evaluates the effective noise power spectral density of a detector at different frequencies. The module contains the LISA sensitivity curve that can be tweaked by adjusting parameters such as the observation time, response function, and the arm length. For the Galactic confusion noise, the module uses the Robert et al. 2019 model.

*Annotations:* 

    """
    Calculates the effective power spectral density for all instruments.
    """
  
*Inputs:*

- {**name:** `f`, **description:** `Frequencies to evaluate sensitivity curve`, **type:** `Float`, **optional:** `False`}
- {**name:** `instrument`, **description:** `Instrument to use`, **default:** `'LISA'`, **type:**`String`, **optional:** `True`}
- {**name:** `custom_psd`, **description:** `Custom function for computing the PSD`, **default:** `None`, **type:** `Funcion`, **optional:** `True`}
- {**name:** `t_obs`, **description:** `Observation time`, **default:** `4`, **type:** `Float`, **optional:** `True`}
- {**name:** `L`, **description:** `LISA arm length in metres`, **default:** `2.5`, **type:** `Float`, **optional:** `True`}
- {**name:** `approximate_R`, **description:** `Whether to approximate response function`, **default:** '`False`', **type:** `Bool`, **optional:** `True`}
- {**name:** `confusion_noise`, **description:** `Galactic confusion noise`, **default:** `'robson19'`, **type:** `Various`, **optional:** `True`}

*Outputs:*

- {**name:** `psd`, **type:** `Float`, **description:** `Effective power strain spectral density`}

*Implementation:*

```python
import GravitationalWaves.psd as psd
import numpy as np
import astropy.units as u

frequency_range = np.logspace(-5, 0, 1000) * u.Hz
PSD = psd.power_spectral_density(f = frequency_range) # get noise amplitude
noise_amplitude = np.sqrt(frequency_range * PSD)
```


</div>




----------------------------------------




#### Module 4. utils.py


<div class = "methods1">

*Name:* utils

*Description:* The utils module is a collection of miscellaneous utility functions for computing conversions between variables, the chirp mass of binaries, and expressions from the Peters equation (Peters and Mathews 1963), such as the relative power of gravitational radiation and enhancement factor.

*Annotations:*  

    """
    Computes chirp mass of binaries
    Compute g(n, e) from Peters and Mathews (1963) Eq.20
    Converts between orbital frequency and semi-major axis using Kepler's third law
    """
  
*Inputs:*

- {**name:** `m_1`, **description:** `Primary mass`, **type:** `Float`,  **optional:** `False`}
- {**name:** `m_2`, **description:** `Secondary mass`, **type:** `Float`,  **optional:** `False`}
- {**name:** `n`, **description:** `Harmonic(s) of interest`, **type:** `Int`,  **optional:** `False`}
- {**name:** `e`, **description:** `Eccentricity`, **type:** `Float`,  **optional:** `False`}
- {**name:** `f_orb`, **description:** `Orbital frequency`, **type:** `Float`,  **optional:** `False`}


*Outputs:*

- {**name:** `m_c`, **type:** `Float`, **description:** `Chirp mass`}
- {**name:** `g`, **type:** `Array`, **description:** `Power of gravitational radiation at nth harmonic`}
- {**name:** `f_orb`, **type:** `Float`, **description:** `Orbital frequency`}
- {**name:** `a`, **type:** `Float`, **description:** `Semi-major axis`}


*Implementation:*

```python
from GravitationalWaves import utils
import astropy.units as u

mass1 = 11 * u.Msun
mass2 = 11 * u.Msun
m_c = utils.chirp_mass(m_1 = mass1, m_2 = mass2)
```

</div>






----------------------------------------



#### Module 5. visualizations.py


<div class = "methods1">

*Name:* visualizations

*Description:* The visualization module uses matplotlib and seaborn to plot quick visuals for analyzing of a collection of sources. The plots include 1- and 2-dimensional distributions and the LISA sensitivity curve.

*Annotations:* 

    """
    Plot a 1D distribution of `x` and `y`
    Plot a 2D distribution of `x` and `y`
    Plot the LISA sensitivity curve
    """
  
*Inputs:*

- {**name:** `x`, **description:** `Variable to plot on the x axis`, 
         **type:** `Float`, **optional:** `False`}
- {**name:** `y`, **description:** `Variable to plot on the y axis`, **default:** `None`, **type:** `Float`, **optional:** `True`}
- {**name:** `weights`, **description:** `Weights for each variable pair`, **type:** `Float`, **optional:** `False`}
- {**name:** `disttype`, **description:** `Type of distribution plot to use`, **type:** `{ "scatter", "kde" }`, **optional:** `False`}
- {**name:** `frequency_range`, **description:** `Frequency values to plot sensitivity curve`, **type:** `Float`, **optional:** `False`}
- {**name:** `y_quantity`, **description:** `Quantity to plot on y axis`, **default:** `None`, **type:** `{ "ASD", "h_c" }`, **optional:** `True`}


*Outputs:*

- {**name:** `fig`, **type:** `matplotlib Figure`, **description:** `Figure of distribution`}
- {**name:** `ax`, **type:** `matplotlib Axis`, **description:** `Axis of distribution plot`}

*Implementation*:


```python
import GravitationalWaves.visualization as vis
import numpy as np
import astropy.units as u

fig, ax = sources.plot_source_variables(xstr = "f_orb", ystr = "snr", disttype = "kde", log_scale = (True, True), show=False, which_sources = sources.snr > 0)
right_ax = ax.twinx()
frequency_range = np.logspace(np.log10(2e-6), np.log10(2e-1), 1000) * u.Hz

vis.plot_sensitivity_curve(frequency_range = frequency_range, fig = fig, ax = right_ax)
```



</div>




----------------------------------------



#### Module 6. wavesim.py


<div class = "methods1">

*Name:* wavesim

*Description:* The gwsim module contains functions for simulating gravitational waves from binary black holes. The procedure includes parts for simulating the inspiral portions, pre-matching and post-matching parts for simulating the merger/ringdown portions, and parts for matching together the inspiral and merger/ringdown portions. 

*Annotations:* 

    """
    Parts of the procedure for simulating the inspiral portions of gravitational waves, collected into modular functions.
    """
  
*Inputs:*

- {**name:** `logMc`, **description:** `log (Mc/Msun)`, **type:** `Float`, **optional:** `False`}
- {**name:** `q`, **description:** `Eccentricity`, **type:** `Float`, **optional:** `False`}        
- {**name:** `flow`, **description:** `Frequency flow`, **default:** `10.0`, **type:** `Float`, **optional:** `True`}    
- {**name:** `merger_type`, **description:** `Binary black hole or neutron star merger`, **type:** `String`, **optional:** `False`}
- {**name:** `D`, **description:** `Distance`, **default:** `100.0`, **type:** `Float`, **optional:** `True`}
         
*Outputs:*

- {**name:** `i_m_amp`, **type:** `Float`, **description:** `Strain envelope amplitude`}
- {**name:** `i_m_freq`, **type:** `Float`, **description:** `Frequency over time for whole waveform`}


</div>



----------------------------------------


## Functional Specification

### Usage Examples


With this package, users can


1. Simulate a gravitational wave signal from a binary merger of two black holes, and output the frequency and strain amplitude data for that signal.
2. Create a random collection of possible LISA binary sources and calculate their strain amplitudes given any range of frequency harmonics.
3. Calculate and plot the sensitivity curve of a gravitational wave detector
4. Check  whether a binary source is eccentric/circular and evolving/stationary
5. Compute the signal-to-noise ratio (SNR) of a single source binary system, as well as a collection of systems
6. Plot either 1D or 2D parameter distributions for a collection of sources
7. Determine the number of detectable binaries from a population of sources



\ 


### Demo 1. Single source SNR calculation


The most basic use case for GravitationalWaves is to calculate the signal-to-noise ratio of a single stellar-mass binary system. We do this by using the package's source module to generate a Source class.  First we must import the package module. As soon as we use import statements, we can use modules, which can be built-in, installed third-party modules, or internal project modules. The best import statement is simply `import modu`. Using `from modu import func` is a way to pinpoint the function to import and put it in the local namespace.


To import the file `source.py` in a directory, we use the statement `import GravitationalWaves.source`. This statement looks in the directory for `__init__.py` and then for `source.py`, executing all top-level statements in both files. Consequently, any variable, function, or class defined in `source.py` is available in the local namespace. 




```python
import GravitationalWaves
import GravitationalWaves.source 
import astropy.units as u

source1 = source.Source(m_1 = 11 * u.Msun,
                        m_2 = 11 * u.Msun,
                        ecc = 0.3,
                        f_orb = 1e-4 * u.Hz,
                        dist = 9 * u.kpc,
                        interpolate_g = False)
                          
source1.get_snr()
```

For this example, GravitationalWaves checks whether the source is eccentric/circular and evolving/stationary, and chooses the fastest method to accurately calculate the SNR. 


----------------------------------------------


### Demo 2. Multiple source SNR calculate

In the next example, we use GravitationalWaves to calculate the detectability of a collection of sources. We first import the package, then create a random set of possible LISA sources.


```python
import legwork.source as source
import numpy as np
import astropy.units as u

# create a random collection of sources
nVals = 1800
mass1 = np.random.uniform(0, 12, n_values) * u.Msun
mass2 = np.random.uniform(0, 12, n_values) * u.Msun
eccInit = 1 - np.random.power(6, n_values)
lumDist = np.random.normal(9, 1.5, n_values) * u.kpc
orbFreq = 12**(-6 * np.random.power(3, n_values)) * u.Hz
```


Using these random sources, we can instantiate a Source class.

```python
sources = source.Source(m_1 = mass1, m_2 = mass2, ecc = eccInit,  dist = lumDist, f_orb = orbFreq)
```

Now, we can calculate the SNR (signal-to-noise ratio) for these sources. This function splits the sources based on whether they are stationary/evolving and circular/eccentric.


```python
snr = sources.get_snr(verbose = True)
```

The SNR values are now stored in `sources.snr`, and we can mask any values that don’t meet some detectable threshold. In the following, we set the threshold to 7.


```python
detectableSources = sources.snr > 7
print("{} out of the {} sources are detectable".format(
    len(sources.snr[detectableSources]), nVals))
```






----------------------------------------



## Discussion

### Software Design Decisions

Overall, this project taught me how to structure a Python project. Structuring GravitationalWaves means making decisions about how the project can best achieve its goals: simulating, analyzing, and detecting gravitational waves. In other words, the structure requires writing clean code, with clear logic and dependencies, and organizing files and folders in the file system. Writing clean code with clear logic is part of the architectural task of building out the different components of a project and their interactions. Organizing a project requires making decisions, such as which features should go into a module, how data should flow through the project, and which features and functionality should be grouped together. 



By structuring the Python project, I learned how to divide the code base into clean, efficient modules using packages. This step requires building modules as an abstraction layer to divide the code into sections containing related data and functionality. For example, this project separates the layers that deal with plotting results from the layers that deal with miscellaneous utility functions by grouping all plotting functions in the visualization module and all utility functions in the utils module. Overall, the straightforward module importing model and packaging system provided in Python makes it notably easier to structure a Python package. 




#### Design Considerations:

- Simplifying the number of parameter given to a gravitational wave source model to provide easier use for users.
- Making the analysis and calculations of gravitational wave properties more specific and targeted.
- Combining the features of the `legwork` and `riroriro` package so that users can both simulate theoretical gravitational wave forms and analyze existing sources.




\ 



### Software Extensibility



Overall, GravitationalWaves is designed to help calculate binary sources of gravitational waves, whether simulated or LISA-like. This package aims to provide a way to study and better understand the detectability of such compact object binaries. Future work includes adding more functions, equations, and modules to the package to implement and analyze gravitational-wave emission, gravitational wave strain, SNR, and visualization modules, and see these implementations' effect on orbital evolution. This will require adding more modules and functions to compute higher-level linear algebra-based equations and mathematical models.




--------------------------------------------



## References 


